Nonparametric Clustering from Energy Statistics

### 2017-08-13 qcqp_clustering/draft/energy_clustering.tex
more polishing

\begin{abstract}
Energy statistics 
was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics,
and it provides a nonparametric test for equality of distributions.
It was generalized to 
metric spaces of strong negative type,
and more recently a connection with reproducing kernel Hilbert spaces (RKHS)
was proposed.
Although  
extensively used in statistics, it
has not been previously incorporated in machine learning problems.
Here we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP), which we show
to be equivalent
to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of 
kernel $k$-means from energy statistics.
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, which makes connection to 
graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative
algorithm based on Hartigan's method, which in this case
has the same computational cost 
as kernel $k$-means algorithm but usually with better clustering
performance. We provide carefully designed numerical experiments
showing the superiority of this method compared to
kernel $k$-means, standard $k$-means, and gaussian mixture models.
\end{abstract}

@@

Energy statistics \cite{Szkely2013}
is based on a 
notion of statistical potential energy between probability distributions,
in close analogy to Newton's gravitational
potential in classical mechanics. 
It provides a nonparametric test for equality of distribution which
is achieved under minimum energy. When probability distributions
are different, this statistical potential 
energy diverges as sample size increases.
Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, 
multi-sample tests of equality of distributions, analysis of variance
\cite{RizzoVariance}, nonlinear
dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and  
hierarchical clustering \cite{RizzoClustering} by extending Ward's
method of minimum variance. (We refer to \cite{Szkely2013} and 
references therein for an overview of energy statistics.)
Moreover, an energy statistics formulation
to clustering was already proposed in \cite{Kgroups} 
which greatly motivated this paper.

@@

The clustering problem has a long history in machine learning.
The most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of assigning a data point to
the cluster with closest center. The only statistical 
information about each cluster
comes from its mean and thus it is sensitive to outliers. Moreover, $k$-means
is strongly parametric, assuming that data is isotropically and normally
distributed since it is a special case of gaussian mixture models (GMM) 
through the expectation maximization (EM) algorithm. Nevertheless,
$k$-means works very well when data is linearly separable 
in Euclidean space, with a clear resolution of cluster centers and without
significant overlap of the points belonging to different clusters.

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert
space by using its inner product. This enables one to cluster way more
complicated datasets, however, the kernel choice remains the biggest challenge
since there is no principled theory behind this. The well-known
kernel $k$-means optimization problem is nothing but $k$-means in the
feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd} of
using the mean of each cluster in the feature space. 
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics was considered
in \cite{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection to kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the
principle behind clustering is to maximize the statistical energy, 
enforcing probability distributions associated to each cluster
to be different from one another.
We provide a precise mathematical formulation to this
statement, showing that there is only one possible optimization problem
consistent with energy statistics. Moreover, we show that this is
a quadratically constrained
quadratic program (QCQP) in the associated RKHS. Our results immediately 
establish the connection 
to kernel methods used in machine learning, 
by showing that this QCQP is equivalent
to kernel $k$-means optimization problem. 
The equivalence between kernel $k$-means, 
spectral clustering, and graph partitioning problems was already established
in \cite{Dhillon,Dhillon2}. We show how these connections arise
from a weighted version of energy statistics.

Our algorithmic contribution is to use Hartigan's method
\cite{Hartigan} to find local solutions of the above mentioned QCQP.
This approach was also proposed in \cite{Kgroups}. In this case,
the resulting algorithm has the same
time complexity as kernel $k$-means algorithm, however, the numerical 
results provide compelling evidence that this method
is more accurate and robust than kernel $k$-means. This evidence
is also supported by general theoretical results \cite{Telgarsky,Slonin}.
Moreover, in our experiments we put in
evidence the nonparametric aspect of energy
statistics based clustering, which provides a default kernel, showing that
it is able to perform accurately on data coming very different 
distributions, contrary to $k$-means and GMM for instance.

@@

There is an equivalence between energy distance, 
commonly used in statistics,
and distances between embeddings of distributions in 
RKHS, commonly used in machine learning. 
This equivalence was established
in \cite{Sejdinovic2013}. Let us first recall the definition of
RKHS. Let $\HH$ be a Hilbert space of real-valued functions
over $\mathcal{X}$. A function 
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$ is a reproducing kernel of $\HH$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
for all $x \in \mathcal{X}$.
\item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
all $x\in\mathcal{X}$ and $f\in \HH$.
\end{enumerate}
In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$,
there is a unique 
$\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
of $\HH$.
If such a \emph{kernel} 
function $\kk$ exists, then $\HH$ is called a RKHS. The above two 
properties immediately imply that $\kk$ is symmetric and positive
definite. Indeed, notice that
$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and by definition
$\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle$, but
since the inner product is real we have 
$\langle \kkk_y, \kkk_x \rangle = \langle \kkk_x, \kkk_y \rangle$, or
equivalently 
$\kk(y,x) = \kk(x,y)$. Moreover, for any $w \in
\HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$, where
$\{ \kkk_{x_i} \}_{i=1}^n$ is a basis of $\HH$. It follows that
$\langle w, w \rangle_{\HH}  = \sum_{i,j=1}^n c_i c_j \kk(x_i,x_j) \ge 0$,
showing that the kernel is positive definite. If $G$ is a matrix with
elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G$ being
positive semidefinite, i.e. $v^\top G \, v \ge 0$ for any vector
$v \in \mathbb{R}^n$.

@@

The clustering problem \eqref{eq:qcqp2} based on energy statistics
is valid for data living in an \emph{arbitrary} space of negative type where
a semimetric $\rho$, and thus the kernel \eqref{eq:kernel_semimetric}, are
assumed to be known. This clustering method is
\emph{nonparametric} since it does not make any assumptions
about  the distribution of the data, 
contrary to $k$-means and gaussian mixture models (GMM), for example.
%Moreover, this approach \emph{does not} require the concept of the 
%\emph{cluster mean},
%which can be ill-defined for some types of data, such as images for
%instance. Since no mean is involved, the method should also be 
%robust to outliers.
%This method is quite general and makes very few
%assumptions about the data. 
If one uses the standard energy distance
defined in \eqref{eq:energy} 
we have $\rho(x,y) = \| x - y\|$ and this fixes the kernel
through \eqref{eq:kernel_semimetric}. In the same way, 
for data living in a more general
metric space $(\mathcal{X}, \rho)$ the corresponding semimetric $\rho$ fixes
the kernel.
In practice, however,
the clustering quality strongly depend on the choice of a suitable
$\rho$ which measures the similarity between data points,
and is equivalent to choosing an appropriate kernel.
The energy distance \eqref{eq:energy2} fixes a family of choices
$\rho(x,y) = \| x-y\|^\alpha$ with $0 < \alpha\le 2$. 
Nevertheless, if prior information
is available one may choose a more appropriate $\rho$.

@@

\begin{proposition}
\label{th:kernel_kmeans}
For a fixed kernel,
the clustering optimization problem
\eqref{eq:minimize} based on energy statistics 
is equivalent to the kernel $k$-means optimization problem
\eqref{eq:kernel_kmeans}, and both are equivalent to \eqref{eq:qcqp2}.
\end{proposition}
\begin{proof}
Notice that $\| \varphi(x) - \varphi(\mu_j) \|^2 = \varphi(x)^\top \varphi(x)
- 2 \varphi(x)^\top \varphi(\mu_j) + \varphi(\mu_j)^\top \varphi(\mu_j)$,
therefore
\begin{equation}
\label{eq:J}
J = \sum_{j=1}^k \sum_{x\in\C_j} \bigg(
\kk(x,x) - 
\dfrac{2}{n_j} \sum_{y\in \C_j} \kk(x,y) + \dfrac{1}{n_j^2}
\sum_{y,z \in \C_j} \kk(y,z) \bigg).
\end{equation}
The first term is global so it does not contribute to the optimization
problem. Notice that the third term gives
$\sum_{x\in\C_j} \tfrac{1}{n_j^2} \sum_{y,z\in\C_j} \kk(y,z) =
\tfrac{1}{n_j}\sum_{y,z\in\C_j} \kk(y,z)$, which is the same as
the second term. Thus, problem
\eqref{eq:kernel_kmeans} is equivalent to
\begin{equation}
\min_{\C_1,\dotsc,\C_k} \bigg\{ J(\C_1,\dotsc,\C_k) = \max_{\C_1,\dotsc,\C_k}
\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in\C_j} \kk(x,y) \bigg\}
\end{equation}
which is exactly the same as 
\eqref{eq:max_prob} from the energy statistics formulation. Therefore,
once the kernel $\kk$ is fixed, the function 
$W$ given by \eqref{eq:within} is the same
as $J$ in \eqref{eq:kernel_kmeans}.
The remaining of the proof proceeds as 
already shown in the proof of Proposition~\ref{th:qcqp2}, leading to
the optimization problem in the form \eqref{eq:qcqp2}.
\end{proof}

The above result shows that 
kernel $k$-means problem is actually a consequence
of energy statistics. In this vein, kernel $k$-means problem is part of a
statistical framework where we are maximizing distances between probability
distributions. The kernel function $\kk$ arises from a semimetric $\rho$ of
negative type which defines the energy distance between distributions.

@@

For a two-class problem we can use \eqref{eq:w1d} to cluster the data
through a simple algorithm 
as follows. We first order
the entire dataset, $\mathbb{X} \to \bar{\mathbb{X}}$. Then 
we compute \eqref{eq:w1d} for each possible split of $\bar{\mathbb{X}}$
and pick the point which gives the minimum value of $W$. 
This procedure is described in Algorithm~\ref{algo1d}. 
Notice that
this method does not require any initialization,
however,
it only works for one-dimensional data with Euclidean distance. The total
complexity of the algorithm is $\OO(n\log n + n^2) = \OO(n^2)$.

@@

Assuming the true label matrix $Z$ is available, a direct
measure of how different the estimated matrix $\hat{Z}$ 
is from $Z$, up to label
permutations, is given by
\begin{equation}
\label{eq:accuracy}
\textnormal{accuracy}(\hat{Z}) \equiv \max_\sigma
\dfrac{1}{n}\sum_{i=1}^n\sum_{j=1}^k \hat{Z}_{i \sigma(j)} Z_{ij}
\end{equation}
where $\sigma$ is a permutation
of the $k$ cluster groups. 
The accuracy is always between $[0,1]$, where
$1$ corresponds to all points correctly clustered, and 
$0$ to all points wrongly clustered.
For a two-class problem with equal
number of points in each cluster, the value $1/2$ correspond
to chance.

@@

Let us consider two simple experiments with equal number of points
in each cluster. We plot the accuracy \eqref{eq:accuracy} versus the number
of points in each cluster.
The data is clustered using 
Algorithm~\ref{algo1d}, GMM through EM algorithm, and standard $k$-means 
algorithm. In the last two cases 
we use the initialization from $k$-means++ \cite{Vassilvitskii} 
and we run the algorithms
few times choosing the result with  best objective function value. 
Notice that Algorithm~\ref{algo1d} requires no initialization so we only
run it once.
Moreover, for each case we sample $100$ times
and show the average accuracy with error bars indicating
the standard error.
In  Fig.~\ref{fig:1d}a 
we have data sampled from two normal distributions with equal number of
points in each cluster,
\begin{equation}
\label{eq:two_normal}
x \sim \mathcal{N}\big(\mu_i,\sigma_i^2\big) 
\quad 
\mbox{with $\mu_1 = 0$, $\sigma_1=1$ and
$\mu_2 = 5$, $\sigma_2 = 2$.}
\end{equation}
For these distributions the optimal classification accuracy 
obtained from Bayes error
is $ \approx 0.956$ which is indicated by the dashed line in the plot.
We see that the three methods
perform closely. As expected, GMM has a slight advantage over the other
methods since
it corresponds to the true model of the data. 
Energy statistics performs slightly better
than $k$-means. On the other hand, in Fig~\ref{fig:1d}b
we consider two clusters with lognormal distributions,
\begin{equation}
\label{eq:two_lognormal}
\log x \sim \mathcal{N}\big( \mu_i,\sigma_i^2\big) \quad
\mbox{with
$\mu_1 = 0$,
$\sigma_1 = 0.3$ and
$\mu_2 = -1.5$,
$\sigma_2 = 1.5$.}
\end{equation}
The optimal classification accuracy from Bayes error is $\approx 0.852$.
In this case,
Algorithm~\ref{algo1d} provides a very accurate clustering while
GMM and $k$-means basically cluster at chance.
Sometimes GMM/EM  was unable to estimate the parameters 
thus giving zero accuracy.
These two simple experiments from Fig.\ref{fig:1d} illustrate
how energy statistics clustering is nonparametric, being able
to provide high quality clustering in settings where data comes
from very different distributions.

@@

\begin{figure}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
    \INPUT number of clusters $k$, Gram matrix $G$, initial label
    matrix $Z = Z_0$
    \OUTPUT label matrix $Z$ 
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            have the costs of each cluster, according to \eqref{eq:maxQ}
  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ 
        have the number of points in each cluster, obtained 
        from $D = Z^\top Z$
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmin_{\ell} J^{(\ell)}(x_i)$
            according to \eqref{eq:Jell}, for $\ell=1,2,\dots,k$
        \IF{ $j^\star \ne j$} 
            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and
            $Z_{ij^\star} \leftarrow 1$
            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{kmeans_algo}
Kernel $k$-means algorithm 
to find an approximate solution to \eqref{eq:qcqp2}.
\hspace{\fill}
}
\end{algorithm}
\end{figure}

@@

Both condition above are not satisfied 
by Lloyd's method, and consequently by (kernel) $k$-means algorithm.
The next result indicates that Hartigan's has the potential to
escape local optima of Lloyd's.

@@

This means that Algorithm~\ref{kmeans_algo} cannot
improve on a local optima of Algorithm~\ref{algo}. On the other hand,
Algorithm~\ref{algo} might improve on a local optima of 
Algorithm~\ref{kmeans_algo}. Lloyd's method forms Voronoi partitions,
while Hartigan's method groups data
in regions formed by the intersection of spheres called circlonoi cells.
It can be shown that the circlonoi cells are contained within
a smaller volume of a Voronoi cell, and this excess volume grows
exponentially with the dimension of $\mathcal{X}$ 
\cite[Theorems 2.4 and 3.1]{Telgarsky}. 
Points in this excess volume
force Hartigan's method to iterate, contrary
to Lloyd's method. Therefore, Hartigan's method 
can potentially escape local
optima of Lloyd's method. 
Moreover, this improvement should be more prominent as
dimension increases. Also, the improvement grows as $k$
increases.
The empirical results of \cite{Telgarsky} show that 
an implementation of Hartigan's method has comparable execution time 
as an implementation of
Lloyd's method,
but no explicit complexity was provided. In our case, we showed that both
Algorithms~\ref{kmeans_algo} and \ref{algo} have the same time complexity.

>>>>

\begin{abstract}
Energy statistics 
was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics,
and it provides a nonparametric test for equality of distributions.
It was generalized to 
metric spaces of strong negative type,
and more recently a connection with reproducing kernel Hilbert spaces 
was proposed.
Although  
extensively used in statistics, it
has not been previously incorporated in machine learning problems.
Here we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP), which we show
to be equivalent
to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of 
kernel $k$-means from energy statistics.
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, which makes connection to 
graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative
algorithm based on Hartigan's method, which in this case
has the same computational cost 
as kernel $k$-means algorithm but usually with better clustering
quality. We provide carefully designed numerical experiments
showing the superiority of this method compared to
kernel $k$-means, standard $k$-means, and gaussian mixture models.
\end{abstract}

@@

Energy statistics \cite{Szkely2013}
is based on a 
notion of statistical potential energy between probability distributions,
in close analogy to Newton's gravitational
potential in classical mechanics. 
It provides a nonparametric test for equality of distribution which
is achieved under minimum energy. When probability distributions
are different, this statistical potential 
energy diverges as sample size increases.
Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, 
multi-sample tests of equality of distributions, analysis of variance
\cite{RizzoVariance}, nonlinear
dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and  
hierarchical clustering \cite{RizzoClustering} by extending Ward's
method of minimum variance. We refer to \cite{Szkely2013} and 
references therein for an overview.
Moreover, an energy statistics formulation
to clustering was already proposed in \cite{Kgroups} 
which greatly motivated this paper.

@@

The clustering problem has a long history in machine learning.
The most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of assigning a data point to
the cluster with closest center. The only statistical 
information about each cluster
comes from its mean and it is thus sensitive to outliers. Moreover, $k$-means
is strongly parametric, assuming that data is isotropically and normally
distributed since it is a special case of gaussian mixture models (GMM) 
through the expectation maximization (EM) algorithm. Nevertheless,
$k$-means works very well when data is linearly separable 
in Euclidean space, with a clear resolution of cluster centers and without
significant overlap of the points belonging to different clusters.

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert
space by using its inner product. This enables one to cluster way more
complicated datasets, however, the kernel choice remains the biggest challenge
since there is no principled theory behind this, and it usually introduces
extra parameters that need to be optimized. The well-known
kernel $k$-means optimization problem is nothing but $k$-means in the
feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd} of
using the mean of each cluster in the feature space. 
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics was considered
in \cite{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection to kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the
principle behind clustering is to maximize the statistical energy, 
enforcing probability distributions associated to each cluster
to be different from one another.
We provide a precise mathematical formulation to this
statement, showing that there is only one possible optimization problem
consistent with energy statistics. Moreover, we show that this is
a quadratically constrained
quadratic program (QCQP) in the associated RKHS. Our results immediately 
establish the connection 
to kernel methods used in machine learning, 
by showing that this QCQP is equivalent
to kernel $k$-means optimization problem. 
The equivalence between kernel $k$-means, 
spectral clustering, and graph partitioning problems was already established
\cite{Dhillon,Dhillon2}. We show how these connections arise
from a weighted version of energy statistics.

Our algorithmic contribution is to use Hartigan's method
\cite{Hartigan} to find local 
solutions of the above mentioned QCQP.
This approach was also proposed in \cite{Kgroups}.
In this case,
the resulting algorithm has the same
time complexity as kernel $k$-means algorithm, however, the numerical 
results provide compelling evidence that this method
is more accurate and robust than kernel $k$-means. This evidence
is also supported by general theoretical results \cite{Telgarsky,Slonin}.
Moreover, in our experiments we put in
evidence the nonparametric aspect of energy
statistics based clustering, which provides a family of 
default kernels, showing that
it is able to perform accurately on datasets coming very different 
distributions, contrary to $k$-means and GMM for instance.

@@

There is an equivalence between energy distance, 
commonly used in statistics,
and distances between embeddings of distributions in 
RKHS, commonly used in machine learning. 
This equivalence was established
in \cite{Sejdinovic2013}. Let us first recall the definition of
RKHS. Let $\HH$ be a Hilbert space of real-valued functions
over $\mathcal{X}$. A function 
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$ is a reproducing kernel of $\HH$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
for all $x \in \mathcal{X}$.
\item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
all $x\in\mathcal{X}$ and $f\in \HH$.
\end{enumerate}
In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$,
there is a unique 
$\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
of $\HH$.
If such a \emph{kernel} 
function $\kk$ exists, then $\HH$ is called a RKHS. The above two 
properties immediately imply that $\kk$ is symmetric and positive
definite. Indeed, notice that
$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and by 
definition 
$\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle$, but
since the inner product is real we have 
$\langle \kkk_y, \kkk_x \rangle = \langle \kkk_x, \kkk_y \rangle$, or
equivalently 
$\kk(y,x) = \kk(x,y)$. Moreover, for any $w \in
\HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$ where
$\{ \kkk_{x_i} \}_{i=1}^n$ is a basis of $\HH$. It follows that
$\langle w, w \rangle_{\HH}  = \sum_{i,j=1}^n c_i c_j \kk(x_i,x_j) \ge 0$,
showing that the kernel is positive definite. If $G$ is a matrix with
elements $G_{ij} = \kk(x_i,x_j)$ this is equivalent to $G$ being
positive semidefinite, i.e. $v^\top G \, v \ge 0$ for any vector
$v \in \mathbb{R}^n$.

@@

The clustering problem \eqref{eq:qcqp2} based on energy statistics
is valid for data living in an \emph{arbitrary} space of negative type where
a semimetric $\rho$, and thus the kernel \eqref{eq:kernel_semimetric}, are
assumed to be known. This clustering method is
\emph{nonparametric} since it does not make any assumptions
about  the distribution of the data, 
contrary to $k$-means and GMM, for example.
If one uses the standard energy distance
defined in \eqref{eq:energy} 
we have $\rho(x,y) = \| x - y\|$ and this fixes the kernel
through \eqref{eq:kernel_semimetric}. In the same way, 
for data living in a more general
metric space $(\mathcal{X}, \rho)$ the corresponding semimetric $\rho$ fixes
the kernel.
In practice, however,
the clustering quality strongly depend on the choice of a suitable
$\rho$ which measures the similarity between data points,
and is equivalent to choosing an appropriate kernel.
The energy distance \eqref{eq:energy2} fixes a family of choices
$\rho(x,y) = \| x-y\|^\alpha$ with $0 < \alpha\le 2$. 
Nevertheless, if prior information
is available one may choose a more appropriate $\rho$.

@@

\begin{proposition}
\label{th:kernel_kmeans}
For a fixed kernel,
the clustering optimization problem
\eqref{eq:minimize} based on energy statistics 
is equivalent to the kernel $k$-means optimization problem
\eqref{eq:kernel_kmeans}, and both are equivalent to \eqref{eq:qcqp2}.
\end{proposition}
\begin{proof}
Notice that $\| \varphi(x) - \varphi(\mu_j) \|^2 = \varphi(x)^\top \varphi(x)
- 2 \varphi(x)^\top \varphi(\mu_j) + \varphi(\mu_j)^\top \varphi(\mu_j)$,
therefore
\begin{equation}
\label{eq:J}
J = \sum_{j=1}^k \sum_{x\in\C_j} \bigg(
\kk(x,x) - 
\dfrac{2}{n_j} \sum_{y\in \C_j} \kk(x,y) + \dfrac{1}{n_j^2}
\sum_{y,z \in \C_j} \kk(y,z) \bigg).
\end{equation}
The first term is global so it does not contribute to the optimization
problem. Notice that the third term gives
$\sum_{x\in\C_j} \tfrac{1}{n_j^2} \sum_{y,z\in\C_j} \kk(y,z) =
\tfrac{1}{n_j}\sum_{y,z\in\C_j} \kk(y,z)$, which is the same as
the second term. Thus, problem
\eqref{eq:kernel_kmeans} is equivalent to
\begin{equation}
\min_{\C_1,\dotsc,\C_k} \bigg\{ J(\C_1,\dotsc,\C_k) = \max_{\C_1,\dotsc,\C_k}
\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in\C_j} \kk(x,y) \bigg\}
\end{equation}
which is exactly the same as 
\eqref{eq:max_prob} from the energy statistics formulation. Therefore,
once the kernel $\kk$ is fixed, the function 
$W$ given by \eqref{eq:within} is the same
as $J$ in \eqref{eq:kernel_kmeans}.
The remaining of the proof proceeds as 
already shown in the proof of Proposition~\ref{th:qcqp2}, leading to
\eqref{eq:qcqp2}.
\end{proof}

The above result shows that 
kernel $k$-means problem is actually a consequence
of energy statistics. In this vein, kernel $k$-means is part of a
statistical framework where distances between probability
distributions are maximized. 
The kernel function $\kk$ arises from a semimetric $\rho$ of
negative type which defines the energy distance between distributions.

@@

For a two-class problem we can use \eqref{eq:w1d} to cluster the data
through a simple algorithm 
as follows. We first order
the entire dataset, $\mathbb{X} \to \bar{\mathbb{X}}$. Then 
we compute \eqref{eq:w1d} for each possible split of $\bar{\mathbb{X}}$
and pick the point which gives the minimum value of $W$. 
This procedure is described in Algorithm~\ref{algo1d}. 
Note that
this method does not require any initialization,
however,
it only works for one-dimensional data with Euclidean distance. The total
complexity of the algorithm is $\OO(n\log n + n^2) = \OO(n^2)$.

@@

Assuming the true label matrix $Z$ is available, a direct
measure of how different the estimated matrix $\hat{Z}$ 
is from $Z$, up to label
permutations, is given by
\begin{equation}
\label{eq:accuracy}
\textnormal{accuracy}(\hat{Z}) \equiv \max_\sigma
\dfrac{1}{n}\sum_{i=1}^n\sum_{j=1}^k \hat{Z}_{i \sigma(j)} Z_{ij}
\end{equation}
where $\sigma$ is a permutation
of the $k$ cluster groups. 
The accuracy is always between $[0,1]$, where
$1$ corresponds to all points correctly clustered, and 
$0$ to all points wrongly clustered.
For a two-class problem with balanced
clusters, the value $1/2$ correspond
to chance.

@@

Let us consider two simple experiments with equal number of points
in each cluster. We plot the accuracy \eqref{eq:accuracy} versus the number
of points in each cluster.
The data is clustered using 
Algorithm~\ref{algo1d}, GMM through EM algorithm, and standard $k$-means 
algorithm. In the last two cases 
we use the initialization from $k$-means++ \cite{Vassilvitskii} 
and we run the algorithms
few times choosing the result with  best objective function value. 
Notice that Algorithm~\ref{algo1d} requires no initialization so we only
run it once.
Moreover, for each case we sample $100$ times
and show the average accuracy with error bars indicating
the standard error.
In  Fig.~\ref{fig:1d}a 
we have data sampled from two normal distributions with equal number of
points in each cluster,
\begin{equation}
\label{eq:two_normal}
x \sim \mathcal{N}\big(\mu_i,\sigma_i^2\big) 
\quad 
\mbox{with $\mu_1 = 0$, $\sigma_1=1$ and
$\mu_2 = 5$, $\sigma_2 = 2$.}
\end{equation}
For these distributions the optimal accuracy 
obtained from Bayes classification error
is $ \approx 0.956$ which is indicated by the dashed line in the plot.
We see that the three methods
perform closely. As expected, GMM has a slight advantage over the other
methods since
it corresponds to the true model of the data. 
Energy statistics performs slightly better
than $k$-means. On the other hand, in Fig.~\ref{fig:1d}b
we consider two clusters with lognormal distributions,
\begin{equation}
\label{eq:two_lognormal}
\log x \sim \mathcal{N}\big( \mu_i,\sigma_i^2\big) \quad
\mbox{with
$\mu_1 = 0$,
$\sigma_1 = 0.3$ and
$\mu_2 = -1.5$,
$\sigma_2 = 1.5$.}
\end{equation}
The optimal classification accuracy from Bayes error is $\approx 0.852$.
In this case,
Algorithm~\ref{algo1d} provides a very accurate clustering while
GMM and $k$-means basically cluster at chance.
Sometimes GMM/EM  was unable to estimate the parameters 
thus giving zero accuracy.
The two simple experiments of Fig.~\ref{fig:1d} illustrate
how energy statistics clustering is nonparametric, being able
to provide high quality clustering in settings where data comes
from very different distributions.

@@

\begin{figure}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
    \INPUT number of clusters $k$, Gram matrix $G$, initial label
    matrix $Z = Z_0$
    \OUTPUT label matrix $Z$ 
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            have the costs of each cluster, according to \eqref{eq:maxQ}
  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ 
        have the number of points in each cluster, obtained 
        from $D = Z^\top Z$
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmin_{\ell} J^{(\ell)}(x_i)$
            according to \eqref{eq:Jell}, for $\ell=1,2,\dots,k$
        \IF{ $j^\star \ne j$} 
            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and
            $Z_{ij^\star} \leftarrow 1$
            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{kmeans_algo}
Kernel $k$-means algorithm 
to find a local solution to \eqref{eq:qcqp2}.
\hspace{\fill}
}
\end{algorithm}
\end{figure}

@@

Both conditions of the  above theorem are not satisfied 
by Lloyd's method, and consequently by (kernel) $k$-means algorithm.
The next result indicates that Hartigan's can potentially 
escape local optima of Lloyd's method.

@@

This means that Algorithm~\ref{kmeans_algo} cannot
improve on a local optima of Algorithm~\ref{algo}. On the other hand,
Algorithm~\ref{algo} might improve on a local optima of 
Algorithm~\ref{kmeans_algo}. Lloyd's method forms Voronoi partitions,
while Hartigan's method groups data
in regions formed by the intersection of spheres called circlonoi cells.
It can be shown that the circlonoi cells are contained within
a smaller volume of a Voronoi cell, and this excess volume grows
exponentially with the dimension of $\mathcal{X}$ 
\cite[Theorems 2.4 and 3.1]{Telgarsky}. 
Points in this excess volume
force Hartigan's method to iterate, contrary
to Lloyd's method. Therefore, Hartigan's method 
can escape local
optima of Lloyd's method. 
Moreover, this improvement should be more prominent as
dimension increases. Also, the improvement grows as $k$
increases.
The empirical results of \cite{Telgarsky} show that 
an implementation of Hartigan's method has comparable execution time 
as an implementation of
Lloyd's method,
but no explicit complexity was provided. In our case, we showed that both
Algorithms~\ref{kmeans_algo} and \ref{algo} have the same time complexity.

In \cite{Slonin} Hartigan's method was applied to $k$-means problem
with any Bregman divergence. They showed that the number of Hartigan's
local minima is upper bounded by $\mathcal{O}(1/k)$ 
\cite[Proposition 5.1]{Slonin}. 
In addition, they provide examples where
\emph{any} initial partition correspond to a local optima of Lloyd's 
method, while  the number of local optima in Hartigan's method is small and 
correspond to true partitions of the data. Empirically, the number of
Hartigan's local optima was considerably smaller than the number of Lloyd's
local optima.
These results indicate that Hartigan's method
provides several advantages over Lloyd's method. 
We will see this explicitly in the following 
numerical results where Algorithm~\ref{algo}
outperforms Algorithm~\ref{kmeans_algo} when using the same kernel.
